Designing an effective intervention of motivational interview using sequence classification methods

### 2017-02-28 paper/AMIA/main.tex
update part of abstract and methods sections

\noindent{\bf Abstract}
\textit{Coming soon!}

\section*{Introduction}

In the recent times, more and more data from various domains being produced in the form of event sequences. Thus, sequence classification has become an important task in sequence mining. Sequences can contain discrete (e.g., symbolic sequences such as text, proteins, or DNA) or continuous events (e..g, time series such as ECGs or stocks), depending on the event types. Usually, sequences don't have explicit features and suffer with high dimension of feature space. Even the sequential nature of features is too difficult to capture, which make sequence classification, a more challenging task than feature-based classifications. The sequence classification methods can be divided into three large categories: feature-based classification, distance based method and model based classifications. In feature-based classification, a sequence is transformed into a single vector of features. A supervised learning algorithm such as an support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier. An improvement of feature-based classifier such as shapelet \cite{ye2009time} or pattern-based \cite{kudenko1998feature, lesh1999mining} approach is used for sequence classifications. On the other hand, distance based method measure the similarity between sequences to determine the quality of the classification significance. The most commonly used distance function is euclidian distance \cite{keogh2003need}, while Dynamic Time Wrapping \cite{keogh2000scaling} is used when more flexible matching is desired in time series data. The third category of sequence classification is model based classifiers such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM) or other statistical models. Recently, a hierarchical approach \cite{nallam2016effective} is proposed for sequence classification.

@@

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue. A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exibiting cues. Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns. Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification. In addition, a k-order Markov Model is applied on genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.  Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.

\section*{Methods}
\subsection*{\textit{Data collection}}
The MYSCOPE \cite{} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}. \\

\begin{table*}[htb]
\caption{\textbf{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}}    
\label{table:06}
\centering
\begin{tabular}{lp{3.6cm}lp{8cm}}
\hline
\hline
Annotation  & Description & Speaker & Text \\
\hline
331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do ? mm-hmm meaning is that food available for you ? \\
117 &	Low Uptake, positive	& Adolescent &	Yes \\
301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\
112 &	Change Talk positive	& Adolescent &	my um aunt \\
301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\
112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\
309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\
\hline
\hline
\end{tabular}
\end{table*}

To create the gold standard for our second experiment, we partitioned the adolescent transcripts into sequence of successful and unsuccessful codes which are ended with either negative or positive change talk and commitment language. Table~\ref{table:0001} depicts the distribution of PPC code sequences over the adolescent dataset.

\subsection*{\textit{Sequence classification techniques}}
Generally, a sequence is an ordered list of events. In our project, event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.  Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$. Here, L is fixed and known in advances. In our experiment, known labels are ``successful'' and ``unsuccessful'' motivational interview.

@@

\textbf {Hidden Markov Model}: Hidden Markov Models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.

@@

Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features. \\

@@

Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences. \\

@@

\section*{Conclusion}
In this work, we suggest some successful sequences of codes for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language. We also propose novel features and report the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods for text classification using those features, to help clinical researchers and practitioners assess the feasibility of using these methods for the task of automatic annotation of clinical text using the codebooks of realistic size. We found out that Support Vector Machine using only lexical features consistently outperforms all other classifiers on caregiver and adolescent datasets according to most metrics. Adding contextual and semantic features further improves the performance of SVM on both datasets, achieving close to human accuracy when the codebooks consisting of 16 and 17 classes are used to annotate caregiver and adolescent transcripts, respectively.

This work has important practical implications. First, it can facilitate researchers to establish causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts. Second, since automatic annotation is significantly faster than manual, it can dramatically accelerate the pace of research in behavioral sciences. Third, information that can directly inform and increase the efficiency of clinical practice for a successful interview. Although all experiments were conducted on interview transcripts, the proposed methods and features are not specific to a particular domain of Motivational Interviewing, and thus there is also no prima facie reason to believe that they will not be effective for annotation of any other type of clinical conversation.

Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.

>>>>

\noindent{\bf Abstract}
\textit{Childhood obesity is a serious public health concern in the United States and worldwide. Therefore, there is a need for identification of successful and unsuccessful motivational interviews to facilitate the development of effective interventions for childhood obesity. This paper describes our work on the classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment. Markov chain and hidden Markov models were employed and compared with a different order for the classification of utterance sequences. The classification performances achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using 1st order Markov chain and second order hidden Markov model, respectively. }

\section*{Introduction}
In the recent times, more and more data from various domains being produced in the form of event sequences. Thus, sequence classification has become an important task in sequence mining. Sequences can contain discrete (e.g., symbolic sequences such as text, proteins, or DNA) or continuous events (e..g, time series such as ECGs or stocks), depending on the event types. Usually, sequences don't have explicit features and suffer with high dimension of feature space. Even the sequential nature of features is too difficult to capture, which make sequence classification, a more challenging task than feature-based classifications. The sequence classification methods can be divided into three large categories: feature-based classification, distance based method and model based classifications. In feature-based classification, a sequence is transformed into a single vector of features. A supervised learning algorithm such as an support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier. An improvement of feature-based classifier such as shapelet \cite{ye2009time} or pattern-based \cite{kudenko1998feature, lesh1999mining} approach is used for sequence classifications. On the other hand, distance based method measure the similarity between sequences to determine the quality of the classification significance. The most commonly used distance function is euclidian distance \cite{keogh2003need}, while Dynamic Time Wrapping \cite{keogh2000scaling} is used when more flexible matching is desired in time series data. The third category of sequence classification is model based classifiers such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM) or other statistical models. Recently, a hierarchical approach \cite{nallam2016effective} is proposed for sequence classification.

@@

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequence data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue. A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exibiting cues. Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns. Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification. In addition, a k-order Markov Model is applied on genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.  Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances. A fragment of an adolescent session transcript is presented in Table~\ref{table:anno_examp}. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups. These utterances are then scanned from first to last and divided into successful and unsuccessful sequences when a commitment or change talk is found. A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one dataset that includes all sequences of positive change talk or commitment language (882 sequences) and the other dataset which includes utterances ended with negative change talk or commitment language (192 sequences). In our experiment, these two subsets are utilized to train two Markov chain or HMM.\\

\begin{table*}[htb]
\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
\label{table:anno_examp}
\centering
\begin{tabular}{lp{3.6cm}lp{8cm}}
\hline
\hline
Annotation  & Description & Speaker & Text \\
\hline
331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do ? mm-hmm meaning is that food available for you ? \\
117 &	Low Uptake, positive	& Adolescent &	Yes \\
301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\
112 &	Change Talk positive	& Adolescent &	my um aunt \\
301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\
112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\
309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\
\hline
\hline
\end{tabular}
\end{table*}

\subsection*{\textit{Sequence classification techniques}}
Generally, a sequence is an ordered list of events. In our study, event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.  Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$. Here, L is fixed and known in advances. In our experiment, known labels are ``successful'' and ``unsuccessful'' motivational interview.

@@

\textbf {Hidden Markov Model}: Hidden Markov Models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. HMM (Hidden Markov Model) is a very
powerful tool to statistically model a process that varies in time. It can be seen as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces the time set of observations. A HMM can be fully specified by (1) N, the number of states in the model; (2) M, the number of distinct observation symbols per state; (3) { } A ij = a , the state transition probability distribution; (4) { ( )} B j = b k , the observation symbol probability distribution; and (5) { } Π = π i , the initial state distribution[10]. The Baum-Welch reestimation method was implemented to train a hidden Markov model for each country using the training set. To identify the country of an unknown melody, the Viterbi algorithm was used to decode the sequence and compute its log probabilities respectively using HMM’s trained for different countries. We then assign the melody to the country with the highest log probability.

@@

Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.

@@

Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.

In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the dataset. However, the number of states in higher order Markov chains grows exponentially with the order of the chain. Consequently, it is harder to accurately estimate all the transition probabilities from the training set. That is, there are many high-order states that are not frequently observed in the training set, leading to inaccurate probability estimates. In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain. However, this cannot be easily done for many applications areas. To overcome these problems many variants of the Markov chains have been developed. In the next section we discuss two such variants: the Interpolated Markov Models [SDKW98] and the Selective Markov Models [DK01]. The basic idea of these approaches is to combine portions of different order Markov chains so that to improve the overall classification accuracy.\\

@@

\section*{Conclusion}
It has been shown how Hidden Markov Models could be used to build classifiers for the sequence of behavior codes in motivational interviews. In this work, we suggest some successful sequences of codes for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language. This work can facilitate researchers to establish causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts. Information that can directly inform and increase the efficiency of clinical practice for a successful interview. Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.
